Three-dimensional binary-liquid lattice Boltzmann simulation of microchannels with rectangular cross sections

### 2011-09-25 Text/paper.tex
Fix typos.

\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{color}
\usepackage{tabularx}
\usepackage{overpic}
\usepackage{tikz}
\begin{document}
\begin{frontmatter}
\title{Three-dimensional binary-liquid lattice Boltzmann simulation of microchannels with rectangular
cross sections}
\author[uofa]{A.~Kuzmin\corref{cor1}}
\cortext[cor1]{Corresponding author}
\ead{kuzmin@ualberta.ca}
\author[us]{M.~Januszewski}
\ead{michalj@gmail.com}
\author[schlum]{D.~Eskin}
\ead{deskin@slb.com}
\author[schlum]{F.~Mostowfi}
\ead{fmostowfi@slb.com}
\author[uofa]{J.J.~Derksen}
\ead{jos@ualberta.ca}
\address[uofa]{Chemical and Materials Engineering, University of Alberta\\ 7th Floor, ECERF, 9107
116 St, Edmonton, Alberta, T6G
2V4 Canada}
\address[us]{Insitute of Physics, University of Silesia, 40-007 Katowice, Poland}
\address[schlum]{Schlumberger DBR Technology Center\\ 9450 17 Ave NW, Edmonton, Alberta, T6N 1M9
Canada}
\begin{abstract}
The classical Bretherton problem describes the propagation of gas fingers through liquid media in a
narrow channel with
thin liquid films between bubbles and channel walls. The bubble shape and flow patterns are
complicated functions of the capilary number $Ca$ and Reynolds number $Re$. Recently, we
investigated
an applicability and parameter selection for the two-dimensional Bretherton problem
(flow
between parallel plates) using the free-energy binary liquid lattice Boltzmann method (LBM)
\cite{kuzmin-binary2d}. This paper is the continuation of our previous work with simulations of
three-dimensional channels with rectangular (mostly square) cross sections in the range of the capillary number $0.05\leq
Ca \leq
6.0$.
The flow is driven by a body force, and
periodic boundary conditions are applied in the streamwise direction. The results show that the
binary liquid model is able to correctly capture a number of phenomena occuring in three-dimensional
capillaries, such as the existence of a vortex in front of the bubble and the way bubble radii
depend on the capillary number.  We conclude that lattice Boltzmann free energy binary
liquid model can be used to simulate the Bretherton problem with good accuracy. 
\end{abstract}

@@

For example, the film thickness
is proportional to $Ca^{2/3}$ for small capillary numbers for circular channels
\cite{bretherton,heil-bretherton}. 
Predicting flow patterns and associated mass transfer for Bretheron-type flows
is of significant interest for chemical industry as it is widely used in chemical monolith
microreactors \cite{kreutzer-pressure-drop}. Intensive mass transfer can be achieved because of the
large interfacial area and small diffusion lengths \cite{cerro-bubble-train}. Heat transfer is
also enhanced in comparison with single phase flow \cite{fukugata-levelset}. While it
is possible to calculate the flow analytically for small capillary numbers and nearly zero Reynolds
number \cite{bretherton}, such
results cannot be extrapolated to the wide range of capillary and Reynolds numbers used in the
chemical industry.  Thus, the need of consistent numerical simulations arises.

Flows in two-dimensional geometries (circular tubes, parallel plates) have been studied extensively
in the 
experimental works of \citet{aussillous-deposition}, and \citet{cerro-bubble-train}, and the
numerical works of
\citet{giavedoni-numerical,heil-bretherton}. All abovementioned investigations found that the 
Bretherton
analysis is valid only for small capillary numbers $Ca\leq 0.003$ and deviates from actual
measurements for larger $Ca$.
This is caused by a complex interplay of gravitational,
interfacial, inertial and viscous forces \cite{gupta-review}. Historically, 
\citet{bretherton} neglected inertia effects. \citet{giavedoni-numerical} suggested
that
inertia effects
are negligible for $Ca \leq 0.05$ and have moderate impact for $Ca>0.05$ in the range of Reynolds
number from $0$ to $70$ for two-dimensional flows. Later on, \citet{heil-bretherton}
studied the flow between plates up to $Re=300$. This author noted that while the
influence of inertia on the established film thickness is
insignificant
($7$ percent deviation from the film thickness measured at $Re=0$), the change in Reynolds number
significantly influences  the pressure distrubution and the flow field near the front bubble cap.
\citet{cerro-bubble-train} showed that the mass transfer is strongly affected by flow direction for
small capillary numbers in case of upward and downward flows. Thus, to fully describe the
propagation of the semi-infinite finger in liquid media one needs to take into account the
viscous, gravitational, surface, and inertia forces \cite{gupta-review}.

In comparison with the two-dimensional Bretherton flows (circular tubes, flow between plates), there
is a vast number of experimental
results available for three-dimensional flows in microchannels with square, triangular and
rectangular cross sections. For instance, \citet{cerro-bubble-train} performed a number of
experiments for a
bubble-train flow in capillaries of square cross sections for horizontal, upward and downward flows.
\citet{shikazono-square} obtained
experimental
results for the film thickness dependence on the
capillary number for ethanol/air and water/air mixtures and for square, circular and triangular
shaped mirochannels. They also presented an experimental correlation for  bubble
radii
based on the capillary number $Ca$ and the Weber number $We=Re\,Ca$, where the Reynolds number is
defined as:
\begin{equation}
Re=\frac{U_{\mathrm{bubble}} H_{\mathrm{eff}}}{\nu_{\mathrm{liq}}},
\end{equation}
where $H_{\mathrm{eff}}$ is the channel height and $\nu_{\mathrm{liq}}$ is the kinematic liquid
viscosity.

The experimental works
\cite{shikazono-square,cerro-bubble-train} supported by numerical simulations \cite{heil-threedim,
wang-non-circular} showed interesting phenomena in three-dimensional geometries. It was found
\cite{heil-threedim,wong-films} that for rectangular capillaries there is a transition from an
asymmetric to an axisymmetric shape for a
certain capillary number.
The non-axisymmetric shape describes the wall normal radius (in Fig.
\ref{fig:lbm:benchmark} the wall normal radius is measured along $x$ or $y$ axes) being 
different from the
diagonal radius and the bubble having a non-circular shape in the channel crossection. In this case,
the bubble shape mimics the shape of a rectangle and looks like a rectangle with rounded corners.
Further we adapt the convention ``axial`` for the radius measured along the $x$ or $y$ axes,
see Fig. \ref{fig:lbm:benchmark} \cite{heil-threedim}. The dependence
of the diagonal and axial radii or a square channel on the capillary number is shown in Fig. 
\ref{fig:heil:three:dim}.
\begin{figure}[htb!]
\begin{overpic}[width=\textwidth]{Figures/capillary_width_heil.eps}
\put(50,45){\includegraphics[width=0.4\textwidth]{Figures/inset.eps}}
\put(20,20){\includegraphics[width=0.3\textwidth]{Figures/legend.eps}}
\end{overpic}
%\includegraphics[width=\textwidth]{Figures/capillary_width_heil.eps}
\caption{\citet{heil-threedim} results for the variation of the bubble radii for a range of
capillary numbers for a square channel. One can see the asymmetry between diagonal and axial
diameters for the capillary
numbers $Ca\leq\widehat{Ca}=0.04$. Courtesy of \citet{heil-threedim}. \label{fig:heil:three:dim}}
\end{figure}
The capillary number for square channels at which transition between non-axisymmetric case to
axisymmetric happens is reported in a number of
works ($\widehat{Ca}=0.04$ \cite{cerro-bubble-train},
$\widehat{Ca}=0.1$
\cite{cerro-space,wang-non-circular}, $\widehat{Ca}=0.033$ \cite{heil-threedim}). If the capillary
number is larger
than
the critical capillary number, i.e. $Ca>\widehat{Ca}$, then the bubble becomes axisymmetric with the
radius of the bubble dependent on the capillary number. An example for the bubble radii
dependence on the capillary number is presented in Fig. \ref{fig:heil:three:dim}.

@@

For simplicity, periodic boundary conditions are applied. Therefore  a bubble
train is simulated
instead of a single bubble.  The mutual influence of neighbouring bubbles is minimized
by using a long
channel ($15 H_{\mathrm{eff}}$). Periodic boundary conditions imply as well that the fluid is driven
by a
body force in the framework of the LBM and not with a pressure difference. Therefore, we can not
address simulations with upward and downward flows \cite{cerro-bubble-train} which would require a
simultaneous imposement of pressure boundary conditions and the body force. 
%The pressure boundary
%conditions for the binary liquid model is a challenge. Thus, we consider a flow in the
%horizontal direction only and the upward/downward flows with pressure difference imposed are not
%simulated. 
\begin{figure}[htb!]
\includegraphics*[bb=153 610 410 717,width=\textwidth]{Figures/benchmark_classical.eps} 
\caption{The classical benchmark layout describes a semiinfinite gas bubble
propagation through the liquid media. The inlet pressure is specified $P_{in}$, the outlet is the
free outflow. The bubble propagates with constant velocity $U_{\mathrm{bubble}}$. 
\label{fig:classical:benchmark}}
\end{figure}
\begin{figure}[htb!]
\includegraphics*[bb=152 470 415 713,width=\textwidth]{Figures/benchmark_lbm.eps}
\caption{The lattice Boltzmann benchmark layout as used in present calculations. The dimensions of
the domain are chosen as $H_{\mathrm{eff}}\times H_{\mathrm{eff}}\times 15 H_{\mathrm{eff}}$. The
length of the bubble was chosen as $5 H_{\mathrm{eff}}$ for the film thickness to stabilize.
Periodic boundary conditions are applied in the streamwise direction. The flow is driven by a body
force. \label{fig:lbm:benchmark}}
\end{figure}

@@

% TODO: This paragraph should be revisited, somehow it reads a little weirdly..
It should be noted that the binary liquid model we used has a uniform density. This issue was
partially addressed in the introduction:  to describe the Bretherton problem one needs to take into
account gravitational, viscous, inertial, and surface tension forces. Among all the works, there
is no full 
agreement between results even in the limit of Reynolds number zero. Moreover,
as was indicated
by \citet{kreutzer-taylor} and by \citet{cerro-bubble-train} the pressure distribution, streamlines
and bubble shape are strongly affected by the length of the liquid slug and the length of the
bubble. In the current simulations we minimize the effects of bubbles train and inertia. Inertia
effect are minimized as the largest Reynolds number is not bigger than $10$. The mutual
effects of bubbles motion on each other is minimized by choosing lengths of a bubble and a
channel large enough in comparision with the channel height. However, more thorough studies are
needed in order to understand
precisely the abovementioned effects.

@@

\subsection{Steady-state approach}
\label{sec:steady:state}
We performed different simulations to understand the number of time steps required for the
system to
settle down to a steady state. The grid  was $52 \times 52 \times 750$ which represents
a quarter of the channel with the initial film thickness of $6.5$ lattice units together
with the body force $\frac{\mathrm{d}P}{\mathrm{d}x}=1.6
\cdot 10^{-6}$. The simulation was performed for $250000$ iterations with the step of
$10000$ iterations. The results for time iterations $140000-250000$ are summarised in Table
\ref{table:steady:state}. While the
capillary number variation is around $5$ percent, the radii variation is $0.1$ percent. Note, that
the interface velocity is defined as the bubble tip velocity in the microchannel center. Multiphase
lattice Boltzmann models is known to have spurious currents near the interface
\cite{pooley-spurious,shan-spurious}. Therefore, the bubble interface velocity is influenced by
spurious currents and capillary numbers show larger variation than radii. 
\begin{table}
\begin{tabularx}{\textwidth}{|X|X|X|X|}
\hline
$N_{iter}$&$Ca$&$R_{axis}$&$R_{diag}$\\
\hline
$140000$&$0.1879$&$0.8657$&$0.8708$\\
$150000$&$0.1879$&$0.8652$&$0.8704$\\
$160000$&$0.1878$&$0.8650$&$0.8702$\\
$170000$&$0.1880$&$0.8648$&$0.8701$\\
$180000$&$0.1882$&$0.8647$&$0.8700$\\
$190000$&$0.1885$&$0.8646$&$0.8699$\\
$200000$&$0.1893$&$0.8645$&$0.8698$\\
$210000$&$0.1963$&$0.8645$&$0.8697$\\
$220000$&$0.1925$&$0.8644$&$0.8697$\\
$230000$&$0.1936$&$0.8644$&$0.8697$\\
$240000$&$0.1961$&$0.8645$&$0.8697$\\
$250000$&$0.1932$&$0.8645$&$0.8698$\\
\hline
\end{tabularx}
\caption{Results for the steady-state case. One
can see that $200000$ steps are enough to shape
the bubble and approach the steady state. The small noise in the capillary number is connected to
the identification of the interface and spurious currents in the system. The spurious currents
influence the bubble velocity identification which is taken at the front bubble tip. The binary
liquid model is the continuous interface model. Thus one needs to interpolate phase function data to
obtain the location of the bubble interface $\phi=0$ and its velocity. Thus, there is a larger
deviation for capillary numbers. All other parameters are relaxed to the
steady-state.\label{table:steady:state}}
\end{table}

To further examine the steady-state, the velocity in the streamwise direction is plotted. The values
of the velocity are taken on the contour where $\phi=0$. This corresponds
 to the bubble interface. That allows to check whether the bubble is propagating as a whole rigid
body or it's shrinking or elongating. This is one of the characteristics of the steady-state.  One
can see in Fig. \ref{fig:velocity:contour} the contour plot and the
corressponing streamwise velocity
component. The crossection is a plane $x=0$, where $x$ points
towards the reader, $z$ is the streamwise direction. As fluxes inside the bubble have
clockwise and counterclockwise directions, see Section \ref{sec:velocity}, one needs to thoroughly
examine only the points corresponding to the front and the rear bubble caps. If they are equal, then
it is exactly the velocity with which the bubble propagates in the microchannel as a rigid body.
For a physical analogy, one can imagine the rotating ball propagating in the streamwise direction.
The front and rear ball points have the same streamwise velocity. 
\begin{figure}[ht]
\includegraphics[width=\textwidth]{Figures/velocity_interface_contour.eps}\\
\includegraphics[width=\textwidth]{Figures/velocity_interface_values.eps}\\
\caption{The bubble shape and the corresponding streamwise velocities taken on the bubble
interface after $240000$ iterations. Directions $y$ and $z$ are scaled to the half of the channel
$H_{\mathrm{eff}}/2$. One can see that the bubble front and rear caps are propagating with nearly
the
same velocity, that is the actual bubble velocity. For more details on velocity patterns
see Section \ref{sec:velocity}. \label{fig:velocity:contour}}
\end{figure}

\subsection{Radii transition}
Due to computational power restrictions we can only access the
region of $Ca\geq 0.05$. In this region the proper resolution of the film thickness
is achievable.  To identify the critical capillary number $\widehat{Ca}$ a number of
simulations
were conducted, see Table \ref{table:transition:results}. In comparison with the results of the
\citet{heil-threedim}
($\widehat{Ca}=0.04$) our
results are closer to the VOF continuous interface method simulation by
\citet{wang-non-circular} ($\widehat{Ca}=0.1$). Due to linear approximation we consider a bubble to have a circular cross section if $R_{axis}$ and $R_{diag}$ differ by less than $1\%$. The transition for current simulations
happens for $Ca\approx 0.09$ calculated as linear interpolation between data presented in Table
\ref{table:transition:results}. 
%Our data also fall within $5$ percent radii
%variation on the
%capilllary number of experimental dependency of
%\citet{shikazono-square}.   
One can see
two examples of non-axisymmetric and axisymmetric bubble shapes for $Ca=0.053$ and $Ca=1.13$, see
Fig.
\ref{fig:crosssections:sym}.  
\begin{table}
\begin{tabularx}{\textwidth}{|X|X|X|X|}%X|X|}
\hline
$Ca$&$R_{axis}$&$R_{diag}$&$Re$\\ %&$R_{axis,Wang}$&$R_{diag,Wang}$\\
\hline
$0.053$&$0.95$&$1.01$&$0.907$\\%&$0.99$&$0.99$\\
$0.078$&$0.93$&$0.95$&$1.062$\\%&$0.96$&$0.96$\\
$0.102$&$0.91$&$0.91$&$1.392$\\
$0.132$&$0.89$&$0.89$&$1.80$\\%&$0.93$&$0.93$\\
\hline
\end{tabularx}
\caption{Simulation results in terms of $R_{axis}$ and $R_{diag}$ for the transition region between
non-axisymmetric and axisymmetric cases. The transition occurs at $\widehat{Ca}=0.09$, which is a
linear interpolation between data with radii not equal and equal to each other. The grid for
simulations was chosen as $50\times50\times1500$ for a quarter of a channel. That means that the
corresponding interpolation error of determining radii is a half of the reverse grid number,
$\frac{1}{2 N_x}=0.01$. Thus, if radii values are within $1$ percent we consider them equal to
each other. We also included a corresponding Reynolds number to show that inertia effects can be
neglected.
\label{table:transition:results}}
\end{table}
\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{Figures/phase_crossection_ca5.eps}
\includegraphics[width=0.5\textwidth]{Figures/phase_crossection_ca13.eps}\\
\caption{Crosssections of the phase field in the middle of the bubble for $Ca=0.053$
(left) and $Ca=1.13$ (right). Crosssections are rescaled on the half channel height. Other
parameters are indicated in Table \ref{table:transition:results}. One can see that the left picture
is non-axisymmetric in contrast
with the right one, i.e. $R_{diag}\neq R_{axis}$. The transition happens at
$\widehat{Ca}=0.09$.\label{fig:crosssections:sym}}
\end{figure}

>>>>

\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{color}
\usepackage{tabularx}
\usepackage{overpic}
\usepackage{tikz}
\begin{document}
\begin{frontmatter}
\title{Three-dimensional binary-liquid lattice Boltzmann simulation of microchannels with rectangular
cross sections}
\author[uofa]{A.~Kuzmin\corref{cor1}}
\cortext[cor1]{Corresponding author}
\ead{kuzmin@ualberta.ca}
\author[us]{M.~Januszewski}
\ead{michalj@gmail.com}
\author[schlum]{D.~Eskin}
\ead{deskin@slb.com}
\author[schlum]{F.~Mostowfi}
\ead{fmostowfi@slb.com}
\author[uofa]{J.J.~Derksen}
\ead{jos@ualberta.ca}
\address[uofa]{Chemical and Materials Engineering, University of Alberta\\ 7th Floor, ECERF, 9107
116 St, Edmonton, Alberta, T6G
2V4 Canada}
\address[us]{Insitute of Physics, University of Silesia, 40-007 Katowice, Poland}
\address[schlum]{Schlumberger DBR Technology Center\\ 9450 17 Ave NW, Edmonton, Alberta, T6N 1M9
Canada}
\begin{abstract}
The classical Bretherton problem describes the propagation of gas fingers through liquid media in a
narrow channel with
thin liquid films between bubbles and channel walls. The bubble shape and flow patterns are
complicated functions of the capillary number $Ca$ and Reynolds number $Re$. Recently, we
investigated
an applicability and parameter selection for the two-dimensional Bretherton problem
(flow
between parallel plates) using the free-energy binary liquid lattice Boltzmann method (LBM)
\cite{kuzmin-binary2d}. This paper is the continuation of our previous work with simulations of
three-dimensional channels with rectangular (mostly square) cross sections in the range of the capillary number $0.05\leq
Ca \leq
6.0$.
The flow is driven by a body force, and
periodic boundary conditions are applied in the streamwise direction. The results show that the
binary liquid model is able to correctly capture a number of phenomena occurring in three-dimensional
capillaries, such as the existence of a vortex in front of the bubble and the way bubble radii
depend on the capillary number.  We conclude that lattice Boltzmann free energy binary
liquid model can be used to simulate the Bretherton problem with good accuracy. 
\end{abstract}

@@

For example, the film thickness
is proportional to $Ca^{2/3}$ for small capillary numbers for circular channels
\cite{bretherton,heil-bretherton}. 
Predicting flow patterns and associated mass transfer for Bretherton-type flows
is of significant interest for chemical industry as it is widely used in chemical monolith
microreactors \cite{kreutzer-pressure-drop}. Intensive mass transfer can be achieved because of the
large interfacial area and small diffusion lengths \cite{cerro-bubble-train}. Heat transfer is
also enhanced in comparison with single phase flow \cite{fukugata-levelset}. While it
is possible to calculate the flow analytically for small capillary numbers and nearly zero Reynolds
number \cite{bretherton}, such
results cannot be extrapolated to the wide range of capillary and Reynolds numbers used in the
chemical industry.  Thus, the need of consistent numerical simulations arises.

Flows in two-dimensional geometries (circular tubes, parallel plates) have been studied extensively
in the 
experimental works of \citet{aussillous-deposition}, and \citet{cerro-bubble-train}, and the
numerical works of
\citet{giavedoni-numerical,heil-bretherton}. All abovementioned investigations found that the 
Bretherton
analysis is valid only for small capillary numbers $Ca\leq 0.003$ and deviates from actual
measurements for larger $Ca$.
This is caused by a complex interplay of gravitational,
interfacial, inertial and viscous forces \cite{gupta-review}. Historically, 
\citet{bretherton} neglected inertia effects. \citet{giavedoni-numerical} suggested
that
inertia effects
are negligible for $Ca \leq 0.05$ and have moderate impact for $Ca>0.05$ in the range of Reynolds
number from $0$ to $70$ for two-dimensional flows. Later on, \citet{heil-bretherton}
studied the flow between plates up to $Re=300$. This author noted that while the
influence of inertia on the established film thickness is
insignificant
($7$ percent deviation from the film thickness measured at $Re=0$), the change in Reynolds number
significantly influences  the pressure distribution and the flow field near the front bubble cap.
\citet{cerro-bubble-train} showed that the mass transfer is strongly affected by flow direction for
small capillary numbers in case of upward and downward flows. Thus, to fully describe the
propagation of the semi-infinite finger in liquid media one needs to take into account the
viscous, gravitational, surface, and inertia forces \cite{gupta-review}.

In comparison with the two-dimensional Bretherton flows (circular tubes, flow between plates), there
is a vast number of experimental
results available for three-dimensional flows in microchannels with square, triangular and
rectangular cross sections. For instance, \citet{cerro-bubble-train} performed a number of
experiments for a
bubble-train flow in capillaries of square cross sections for horizontal, upward and downward flows.
\citet{shikazono-square} obtained
experimental
results for the film thickness dependence on the
capillary number for ethanol/air and water/air mixtures and for square, circular and triangular
shaped microchannels. They also presented an experimental correlation for  bubble
radii
based on the capillary number $Ca$ and the Weber number $We=Re\,Ca$, where the Reynolds number is
defined as:
\begin{equation}
Re=\frac{U_{\mathrm{bubble}} H_{\mathrm{eff}}}{\nu_{\mathrm{liq}}},
\end{equation}
where $H_{\mathrm{eff}}$ is the channel height and $\nu_{\mathrm{liq}}$ is the kinematic liquid
viscosity.

The experimental works
\cite{shikazono-square,cerro-bubble-train} supported by numerical simulations \cite{heil-threedim,
wang-non-circular} showed interesting phenomena in three-dimensional geometries. It was found
\cite{heil-threedim,wong-films} that for rectangular capillaries there is a transition from an
asymmetric to an axisymmetric shape for a
certain capillary number.
The non-axisymmetric shape describes the wall normal radius (in Fig.
\ref{fig:lbm:benchmark} the wall normal radius is measured along $x$ or $y$ axes) being 
different from the
diagonal radius and the bubble having a non-circular shape in the channel cross section. In this case,
the bubble shape mimics the shape of a rectangle and looks like a rectangle with rounded corners.
Further we adapt the convention ``axial`` for the radius measured along the $x$ or $y$ axes,
see Fig. \ref{fig:lbm:benchmark} \cite{heil-threedim}. The dependence
of the diagonal and axial radii or a square channel on the capillary number is shown in Fig. 
\ref{fig:heil:three:dim}.
\begin{figure}[htb!]
\begin{overpic}[width=\textwidth]{Figures/capillary_width_heil.eps}
\put(50,45){\includegraphics[width=0.4\textwidth]{Figures/inset.eps}}
\put(20,20){\includegraphics[width=0.3\textwidth]{Figures/legend.eps}}
\end{overpic}
%\includegraphics[width=\textwidth]{Figures/capillary_width_heil.eps}
\caption{\citet{heil-threedim} results for the variation of the bubble radii for a range of
capillary numbers for a square channel. One can see the asymmetry between diagonal and axial
diameters for the capillary
numbers $Ca\leq\widehat{Ca}=0.04$. Courtesy of \citet{heil-threedim}. \label{fig:heil:three:dim}}
\end{figure}
The capillary number for square channels at which transition between non-axisymmetric case to
axisymmetric happens is reported in a number of
works ($\widehat{Ca}=0.04$ \cite{cerro-bubble-train},
$\widehat{Ca}=0.1$
\cite{cerro-space,wang-non-circular}, $\widehat{Ca}=0.033$ \cite{heil-threedim}). If the capillary
number is larger
than
the critical capillary number, i.e. $Ca>\widehat{Ca}$, then the bubble becomes axisymmetric with the
radius of the bubble dependent on the capillary number. An example for the bubble radii
dependence on the capillary number is presented in Fig. \ref{fig:heil:three:dim}.

@@

For simplicity, periodic boundary conditions are applied. Therefore  a bubble
train is simulated
instead of a single bubble.  The mutual influence of neighbouring bubbles is minimized
by using a long
channel ($15 H_{\mathrm{eff}}$). Periodic boundary conditions imply as well that the fluid is driven
by a
body force in the framework of the LBM and not with a pressure difference. Therefore, we can not
address simulations with upward and downward flows \cite{cerro-bubble-train} which would require a
simultaneous imposition of pressure boundary conditions and the body force. 
%The pressure boundary
%conditions for the binary liquid model is a challenge. Thus, we consider a flow in the
%horizontal direction only and the upward/downward flows with pressure difference imposed are not
%simulated. 
\begin{figure}[htb!]
\includegraphics*[bb=153 610 410 717,width=\textwidth]{Figures/benchmark_classical.eps} 
\caption{The classical benchmark layout describes a semi-infinite gas bubble
propagation through the liquid media. The inlet pressure is specified $P_{in}$, the outlet is the
free outflow. The bubble propagates with constant velocity $U_{\mathrm{bubble}}$. 
\label{fig:classical:benchmark}}
\end{figure}
\begin{figure}[htb!]
\includegraphics*[bb=152 470 415 713,width=\textwidth]{Figures/benchmark_lbm.eps}
\caption{The lattice Boltzmann benchmark layout as used in present calculations. The dimensions of
the domain are chosen as $H_{\mathrm{eff}}\times H_{\mathrm{eff}}\times 15 H_{\mathrm{eff}}$. The
length of the bubble was chosen as $5 H_{\mathrm{eff}}$ for the film thickness to stabilize.
Periodic boundary conditions are applied in the streamwise direction. The flow is driven by a body
force. \label{fig:lbm:benchmark}}
\end{figure}

@@

% TODO: This paragraph should be revisited, somehow it reads a little weirdly..
It should be noted that the binary liquid model we used has a uniform density. This issue was
partially addressed in the introduction:  to describe the Bretherton problem one needs to take into
account gravitational, viscous, inertial, and surface tension forces. Among all the works, there
is no full 
agreement between results even in the limit of Reynolds number zero. Moreover,
as was indicated
by \citet{kreutzer-taylor} and by \citet{cerro-bubble-train} the pressure distribution, streamlines
and bubble shape are strongly affected by the length of the liquid slug and the length of the
bubble. In the current simulations we minimize the effects of bubbles train and inertia. Inertia
effect are minimized as the largest Reynolds number is not bigger than $10$. The mutual
effects of bubbles motion on each other is minimized by choosing lengths of a bubble and a
channel large enough in comparison with the channel height. However, more thorough studies are
needed in order to understand
precisely the abovementioned effects.

@@

\subsection{Steady-state approach}
\label{sec:steady:state}
We performed different simulations to understand the number of time steps required for the
system to
settle down to a steady state. The grid  was $52 \times 52 \times 750$ which represents
a quarter of the channel with the initial film thickness of $6.5$ lattice units together
with the body force $\frac{\mathrm{d}P}{\mathrm{d}x}=1.6
\cdot 10^{-6}$. The simulation was performed for $250000$ iterations with the step of
$10000$ iterations. The results for time iterations $140000-250000$ are summarized in Table
\ref{table:steady:state}. While the
capillary number variation is around $5$ percent, the radii variation is $0.1$ percent. Note, that
the interface velocity is defined as the bubble tip velocity in the microchannel center. Multiphase
lattice Boltzmann models is known to have spurious currents near the interface
\cite{pooley-spurious,shan-spurious}. Therefore, the bubble interface velocity is influenced by
spurious currents and capillary numbers show larger variation than radii. 
\begin{table}
\begin{tabularx}{\textwidth}{|X|X|X|X|}
\hline
$N_{iter}$&$Ca$&$R_{axis}$&$R_{diag}$\\
\hline
$140000$&$0.1879$&$0.8657$&$0.8708$\\
$150000$&$0.1879$&$0.8652$&$0.8704$\\
$160000$&$0.1878$&$0.8650$&$0.8702$\\
$170000$&$0.1880$&$0.8648$&$0.8701$\\
$180000$&$0.1882$&$0.8647$&$0.8700$\\
$190000$&$0.1885$&$0.8646$&$0.8699$\\
$200000$&$0.1893$&$0.8645$&$0.8698$\\
$210000$&$0.1963$&$0.8645$&$0.8697$\\
$220000$&$0.1925$&$0.8644$&$0.8697$\\
$230000$&$0.1936$&$0.8644$&$0.8697$\\
$240000$&$0.1961$&$0.8645$&$0.8697$\\
$250000$&$0.1932$&$0.8645$&$0.8698$\\
\hline
\end{tabularx}
\caption{Results for the steady-state case. One
can see that $200000$ steps are enough to shape
the bubble and approach the steady state. The small noise in the capillary number is connected to
the identification of the interface and spurious currents in the system. The spurious currents
influence the bubble velocity identification which is taken at the front bubble tip. The binary
liquid model is the continuous interface model. Thus one needs to interpolate phase function data to
obtain the location of the bubble interface $\phi=0$ and its velocity. Thus, there is a larger
deviation for capillary numbers. All other parameters are relaxed to the
steady-state.\label{table:steady:state}}
\end{table}

To further examine the steady-state, the velocity in the streamwise direction is plotted. The values
of the velocity are taken on the contour where $\phi=0$. This corresponds
 to the bubble interface. That allows to check whether the bubble is propagating as a whole rigid
body or it's shrinking or elongating. This is one of the characteristics of the steady-state.  One
can see in Fig. \ref{fig:velocity:contour} the contour plot and the
corresponding streamwise velocity
component. The cross section is a plane $x=0$, where $x$ points
towards the reader, $z$ is the streamwise direction. As fluxes inside the bubble have
clockwise and counterclockwise directions, see Section \ref{sec:velocity}, one needs to thoroughly
examine only the points corresponding to the front and the rear bubble caps. If they are equal, then
it is exactly the velocity with which the bubble propagates in the microchannel as a rigid body.
For a physical analogy, one can imagine the rotating ball propagating in the streamwise direction.
The front and rear ball points have the same streamwise velocity. 
\begin{figure}[ht]
\includegraphics[width=\textwidth]{Figures/velocity_interface_contour.eps}\\
\includegraphics[width=\textwidth]{Figures/velocity_interface_values.eps}\\
\caption{The bubble shape and the corresponding streamwise velocities taken on the bubble
interface after $240000$ iterations. Directions $y$ and $z$ are scaled to the half of the channel
$H_{\mathrm{eff}}/2$. One can see that the bubble front and rear caps are propagating with nearly
the
same velocity, that is the actual bubble velocity. For more details on velocity patterns
see Section \ref{sec:velocity}. \label{fig:velocity:contour}}
\end{figure}

\subsection{Radii transition}
Due to computational power restrictions we can only access the
region of $Ca\geq 0.05$. In this region the proper resolution of the film thickness
is achievable.  To identify the critical capillary number $\widehat{Ca}$ a number of
simulations
were conducted, see Table \ref{table:transition:results}. In comparison with the results of the
\citet{heil-threedim}
($\widehat{Ca}=0.04$) our
results are closer to the VOF continuous interface method simulation by
\citet{wang-non-circular} ($\widehat{Ca}=0.1$). Due to linear approximation we consider a bubble to have a circular cross section if $R_{axis}$ and $R_{diag}$ differ by less than $1\%$. The transition for current simulations
happens for $Ca\approx 0.09$ calculated as linear interpolation between data presented in Table
\ref{table:transition:results}. 
%Our data also fall within $5$ percent radii
%variation on the
%capilllary number of experimental dependency of
%\citet{shikazono-square}.   
One can see
two examples of non-axisymmetric and axisymmetric bubble shapes for $Ca=0.053$ and $Ca=1.13$, see
Fig.
\ref{fig:crosssections:sym}.  
\begin{table}
\begin{tabularx}{\textwidth}{|X|X|X|X|}%X|X|}
\hline
$Ca$&$R_{axis}$&$R_{diag}$&$Re$\\ %&$R_{axis,Wang}$&$R_{diag,Wang}$\\
\hline
$0.053$&$0.95$&$1.01$&$0.907$\\%&$0.99$&$0.99$\\
$0.078$&$0.93$&$0.95$&$1.062$\\%&$0.96$&$0.96$\\
$0.102$&$0.91$&$0.91$&$1.392$\\
$0.132$&$0.89$&$0.89$&$1.80$\\%&$0.93$&$0.93$\\
\hline
\end{tabularx}
\caption{Simulation results in terms of $R_{axis}$ and $R_{diag}$ for the transition region between
non-axisymmetric and axisymmetric cases. The transition occurs at $\widehat{Ca}=0.09$, which is a
linear interpolation between data with radii not equal and equal to each other. The grid for
simulations was chosen as $50\times50\times1500$ for a quarter of a channel. That means that the
corresponding interpolation error of determining radii is a half of the reverse grid number,
$\frac{1}{2 N_x}=0.01$. Thus, if radii values are within $1$ percent we consider them equal to
each other. We also included a corresponding Reynolds number to show that inertia effects can be
neglected.
\label{table:transition:results}}
\end{table}
\begin{figure}[ht]
\includegraphics[width=0.5\textwidth]{Figures/phase_crossection_ca5.eps}
\includegraphics[width=0.5\textwidth]{Figures/phase_crossection_ca13.eps}\\
\caption{Cross sections of the phase field in the middle of the bubble for $Ca=0.053$
(left) and $Ca=1.13$ (right). Cross sections are rescaled on the half channel height. Other
parameters are indicated in Table \ref{table:transition:results}. One can see that the left picture
is non-axisymmetric in contrast
with the right one, i.e. $R_{diag}\neq R_{axis}$. The transition happens at
$\widehat{Ca}=0.09$.\label{fig:crosssections:sym}}
\end{figure}
